Tracing the cultural evolution of meaning at the semantics-pragmatics interface

### 2016-11-24 paper/paper-v2.tex
draft v2 changes, cleaner (temporary) plot and data files in paper-folder

\definecolor{Red}{RGB}{178,34,34}
\newcommand{\mf}[1]{\textcolor{Red}{[MF: #1]}} 
\usepackage{blkarray}
\usepackage{xspace}

@@

A particularly productive and well studied class of pragmatic enrichments are scalar
implicatures
\citep{horn:1984,Hirschberg1985:A-Theory-of-Sca,LevinsonPragmatics1983,Geurts2010:Quantity-Implic}. Usually,
an utterance of a sentence like ``I own some of Johnny Cash's albums'' will be taken to mean
that the speaker does not own all of them. This is because, if the speaker had them all, he
could have used the stronger word \emph{all} instead of \emph{some} in his utterance and
thereby would have made a more informative statement. Scalar implicatures, especially the
inference from \emph{some} to \emph{some but not all}, have been studied extensively, both
theoretically
\citep[e.g.][]{Sauerland2004:Scalar-Implicat,ChierchiaFox2008:The-Grammatical,Rooyvan-RooijJagerde-Jager2012:Explaining-Quan}
as well as experimentally
\citep[e.g.][]{BottNoveck2004:Some-Utterances,huang+snedeker:2009,GrodnerKlein2010:Some-and-Possib,GoodmanStuhlmuller2013:Knowledge-and-I,DegenTanenhaus2012:Processing-Scal}. While
there is much dispute in this domain about many interesting details, a position endorsed by a
clear majority is that a word like \emph{some} is underspecified to mean \emph{some and maybe
  all} and that the enrichment to \emph{some but not all} is part of some regular enrichment
process with roots in pragmatics.

If this majority view is correct, the question arises how such a division of labor between
semantics and pragmatics could have evolved. Models of meaning evolution abound. There are
simulation-based models studying the evolution of meaning in populations of communicating
agents
\citep{Hurford1989:Biological-Evol,Steels1995:A-Self-Organizi,LenaertsJansen2005:The-Evolutionar,SteelsBelpaeme2005:Coordinating-Pe,BaronchelliPuglisi2008:Cultural-route-,steels:2011,SpikeStadler2016:Minimal-Require}
and there are mathematical models of meaning evolution, mostly coming from in game theory
\citep{lewis:1969,Warneryd1993:Cheap-Talk-Coor,BlumeKim1993:Evolutionary-St,nowak+krakauer:1999,Huttegger2007:Evolution-and-t,Skyrms2010:Signals}. Much
work has focused on explaining basic properties such as compositionality and combinatoriality
\citep[e.g.][]{Batali1998:Computational-S,nowak+krakauer:1999,nowak+etal:2000,KirbyHurford2002:The-Emergence-o,kirby:2002,SmithKirby2003:Iterated-Learni,Gong2007:Language-Evolut,kirby+etal:2015,verhoef+etal:2014,Franke2015:Proto-Syntax}. But
little attention has been paid to the interaction between conventional meaning and pragmatic
use. What is more, many mathematical models explain evolved meaning as a regularity in the
behavior of agents which maps objective states of the world to observable signals. There is no
room in such a purely extensional approach to address the semantics-pragmatics division
directly. We would need to look at richer representations of cognizing agents and their
communicative interaction.

To fill this gap, we here spell out a model of the co-evolution of conventional meaning and
pragmatic reasoning types. The objects of replication and selection are pairs of lexical
meanings and general types of pragmatic behavior, which we represent using state-of-the-art
probabilistic cognitive models of pragmatic language use
\citep{frank+goodman:2012,FrankeJager2015:Probabilistic-p,GoodmanFrank2016:Pragmatic-Langu}. Replication
and selection is described by the \emph{replicator mutator dynamic}, a general and established
model of evolutionary change in large and homogeneous populations
\citep{Hofbauer1985:The-Selection-M,nowak+etal:2000,NowakKomarova2001:Evolution-of-Un,hofbauer+sigmund:2003,Nowak2006:Evolutionary-Dy}. The
approach allows us to study the interaction between (i) evolutionary pressure towards
communicative efficiency and (ii) possible infidelity in the transmission of linguistic
knowledge, such as from inductive learning biases or systematic perceptual errors. Considering
transmission of linguistic knowledge is important because neither semantic meanings nor
pragmatic usage patterns are directly observable. Language learners have to infer these
unobservables from the observable behavior in which they result. We formalize this acquisition
process as a form of Bayesian inference. Our approach thereby contains a well-understood model
of iterated Bayesian learning \citep{griffiths+kalish:2007} as a special case, but combines it
with functional selection, here formalized as the most versatile dynamic from evolutionary game
theory, the replicator dynamic
\citep{TaylorJonker1978:Evolutionary-St}. Section~\ref{sec:model} introduces this model.

Section~\ref{sec:si-case-study} applies this model to a case study on scalar implicatures. We
discuss a setting in which the majority view of underspecified lexical meanings and pragmatic
enrichments emerges if selection and transmission infidelity are combined. In particular, we
show that inductive learning biases of Bayesian learners that favor simpler lexical meanings
can lead to the desired outcome. Additionally, we show that a similar outcome can be achieved
without assuming any cognitive biases, simply as an epiphenomenon of systematic disturbances from
environmental factors. This formal results highlights the frequently overlooked possibility
that channel noise in evolutionary replication can mimic effects of inductive biases.

We see the main contribution of this work as conceptual and technical, not as a definite answer
to the question why scalar implicatures emerged. The work here rather demonstrates how current
probabilistic cognitive modeling of language use and evolutionary modeling can be fruitfully
combined to study the co-evolution of semantics and pragmatics side-by-side. Reversely, the
approach taken here may be seen as a first step towards giving an evolutionary rationale for
empirically successful probabilistic models of language use that embrace the majority view of
the division of labor between semantics and pragmatics. Section~\ref{sec:discussion} elaborates
on these points.

@@

The emergence and change of linguistic structure is influenced by many intertwined
factors. These range from biological and socio-ecological to cultural ones \citep{steels:2011}
\mf{more references?}. Social and ecological pressures determine communicative needs, while
biology determines the architecture that enables and constrains the means by which they can be
fulfilled. In the following, our focus lies on the cultural aspects, wherein processes of
linguistic change are viewed as shaped by language use and its transmission, i.e., as a result
of a process of cultural evolution
\citep{Pagel2009:Human-Language-,ThompsonKirby2016:Culture-Shapes-}.

The idea that language is an adaptation to serve a communicative function has played a pivotal
role in synchronic and diachronic analyses at least since \citeposs{zipf:1949} explanation of
word frequency rankings as a result of competing hearer and speaker preferences (e.g. in
\citealt{martinet:1962, horn:1984,jaeger+vRooij:2007,jaeger:2007,
  piantadosi:2014,kirby+etal:2015}). If processes of selection, such as conditional imitation
or reinforcement, favor more communicatively efficient types of behavior, languages are driven
towards semantic expressivity \citep[e.g.][]{nowak+krakauer:1999,Skyrms2010:Signals}. But
pressure towards communicative efficiency is not the only force that shapes
language. Learnability is another. Natural languages need to be learnable to survive their
faithful transmission across generations. Clearly, an unlearnable code will not make it past
the one happy fellow who invented it. Moreover, even small biases implicit in the acquisition
of a language can built up and have quite striking effects on an evolving language in a process
of iterated learning
\citep{KirbyHurford2002:The-Emergence-o,SmithKirby2003:Iterated-Learni,kirby+etal:2014}. In
sum, natural languages are pressured for expressivity and learnability. But expressivity and
learnability may pull in opposite directions. The opposition becomes particularly clear when
considering the extreme (cf. \citealt{kemp+regier:2012,kirby+etal:2015}). A language with a
single form-meaning association is easy to learn but lacking in expressivity. Conversely, a
language that associates a distinct form with all possible meanings a speaker may want to
convey is maximally expressive but challenging to acquire.

An elegant formal approach to capturing the interaction between expressivity and learnability
is the \emph{replicator mutator dynamic}
\citep{Hofbauer1985:The-Selection-M,nowak+etal:2000,NowakKomarova2001:Evolution-of-Un,hofbauer+sigmund:2003,Nowak2006:Evolutionary-Dy}. In
its simplest, discrete-time formulation, the RMD defines the frequency $x'_i$ of each type $i$
in a population at the next time step as a function of: (i) the frequency $x_i$ of each type
$i$ before the update, (ii) the fitness $f_i$ of each type $i$ before the update, and (iii) the
probability $Q_{ji}$ that an agent who wants to imitate, adopt or learn the type of an agent
with type $j$ ends up acquiring type $i$:
\begin{align}
  \label{eq:RMD_discrete}
  x'_i = \sum_j Q_{ji} \frac{x_jf_j}{\sum_k x_k f_k}\,.
\end{align}
The RMD consists of two components: fitness-based selection and transmission biases. This
becomes most transparent when we consider an equivalent formulation in terms of a step-wise
application of the discrete-time replicator dynamic \citep{TaylorJonker1978:Evolutionary-St}
and subsequent multiplication with a mutation matrix $Q$ of the initial population vector
$\vec{x}$:
\begin{align}
  \label{eq:RMD_discrete_recast}
  x'_i & = (\text{M}(\text{RD}(\vec{x})))_i\,,
\end{align}
where
\begin{align*}
      \left ( \text{RD}(\vec{x}) \right )_i 
         = \frac{x_i f_i}{\sum_k x_k f_k}
 \ \ \ \ \text{and} \ \ \ \ 
  (\text{M}(\vec{x}))_i = (\vec{x} \cdot Q)_i = \left ( \sum_j
          x_j Q_{ji} \right)_i\,.
\end{align*}
If the transmission matrix $Q$ is trivial in the sense that $Q_{ji}=1$ whenever $j=i$, the
dynamic reduces to the replicator dynamic. The replicator dynamic is a model of fitness-based
selection in which the relative frequency of type $i$ will increase with a gradient
proportional to its average fitness in the population. The replicator dynamic is popular and
versatile because it can be derived from many abstract processes of biological and cultural
transmission and selection \citep[for overview and several derivations
see][]{Sandholm2010:Population-Game}, including conditional imitation
\citep[e.g.][]{Helbing1996:A-Stochastic-Be,Schlag1998:Why-Imitate-and} or reinforcement
learning \citep[e.g.][]{BorgersSarin997:Learning-Throug,Beggs2005:On-the-Converge}. If fitness
$f_i$ is the same for all types $i$, the replicator step is the identity map
$ \left ( \text{RD}(\vec{x}) \right )_i = x_i$ and the dynamic reduces to a process of
iteration of the transmission bias encoded in $Q$. In this way, the process in
(\ref{eq:RMD_discrete}), equivalently (\ref{eq:RMD_discrete_recast}), contains a model of
iterated learning \citep{griffiths+kalish:2007}. \mf{should we include a simple example here? I
have an example from a lecture ready at hand; it's a simple coordination game in a
one-population setting.}

@@

Types are what cultural evolution operates on. In standard applications of evolutionary game
theory, types correspond to ways of acting in a game, e.g., either cooperating or defecting in
a prisoner's dilemma. \mf{maybe good to refer back to the example from before if there was
  one?} For our purposes here, types are identified by their cognitive make-up. Since we are
interested in the question under which conditions processes of cultural evolution will favor
specific divisions of labor between lexical meaning and pragmatic use, a type is a pair
consisting of a lexicon and a pragmatic strategy.

Lexica codify the truth-conditions of expressions. A convenient way to represent lexica is by
$(\card{\States}, \card{\Messgs})$-Boolean matrices, where $\States$ is a set of states
(meanings) and $M$ a set of messages (forms available in the language). For example, suppose
that there are two relevant world states $\States = \set{\ssome, \sall}$. In state $\ssome$
Chris owns some but not all of Johnny Cash's albums while in $\sall$ Chris owns them
all. Suppose that there are two messages $\Messgs = \set{\msome, \mall}$ where $\msome$ is
short for a sentence like \emph{Chris owns some of Johnny Cash's albums} and $\mall$ for the
same sentence with \emph{some} replaced by \emph{all}. Lexica for this case would assign a
truth value for each state-message pair. The following two lexica exemplify the distinction
between a lexicalized upper-bound for \emph{some} in $L_4$ and the widely assumed logical
semantics with only a lower-bound in $L_5$. \mf{can we not give these guys more mnemonic
  names?}
\begin{align*}
  L_{4} & = \begin{blockarray}{lcc}
    & \mygray{\msome} & \mygray{\mall} \\
    \begin{block}{l[cc]}
      \mygray{\ssome} & 1 & 0 \\
      \mygray{\sall}  & 0 & 1 \\
    \end{block}
  \end{blockarray} &
  L_{5} & = \begin{blockarray}{lcc}
    & \mygray{\msome} & \mygray{\mall} \\
    \begin{block}{l[cc]}
      \mygray{\ssome} & 1 & 0 \\
      \mygray{\sall}  & 1 & 1 \\
    \end{block}
  \end{blockarray}
\end{align*}

We distinguish between two kinds of pragmatic behavior. {\em Literal interlocutors} produce and
interpret messages literally, being guided only by their lexica. {\em Pragmatic interlocutors}
instead engage in mutual reasoning to inform their choices. Recent probabilistic models of
rational language use
\citep{franke:2009,frank+goodman:2012,FrankeJager2015:Probabilistic-p,GoodmanFrank2016:Pragmatic-Langu}
capture different types of pragmatic behavior in a reasoning hierarchy. The hierarchy's bottom,
level $0$, corresponds to literal language use. Pragmatic language users of level $n + 1$
act (approximately) rational with respect to level-$n$ behavior of their
interlocutors. (\ref{h:level0}) and (\ref{s:level0}) define probabilistic behavior of literal
hearers and speakers respectively: \mf{is it a problem that $S$ denotes speakers and the set of
states?}\mf{Literal speaker definition had $\alpha$ but that has no effect if we have Boolean
semantics; I left this out.}
\begin{flalign}
&H_{0}(s \mid m;L) \propto pr(s) L_{sm} \label{h:level0}\\
&S_{0}(m \mid s;L) \propto \exp(L_{sm}) \label{s:level0}
\end{flalign}
According to (\ref{h:level0}), a literal hearer's interpretation of a message $m$ as a state
$s$ depends on her lexicon and her prior over states, $pr \in \Delta(S)$. For simplicity, in
the following this prior is assumed to be uniform. A literal interpreter with lexicon $L_4$
from above would assign $\ssome$ a probability of $H_0(\ssome \mid \msome; L_4) = 1$ after
hearing $\msome$, while a literal interpreter with lexicon $L_5$ would assign $\ssome$
probability $H_0(\ssome \mid \msome; L_5) = 0.5$. A literal speaker chooses any true message
for a state with equal probability and prefers a true message over a false one. The formulation
in (\ref{s:level0}) allows also false messages to be sent with a small positive probability in
analogy to the definition of pragmatic speakers in probabilistic pragmatic models. This is
necessary to guarantee a mutation matrix with only positive entries (see below). A literal
speaker with lexicon $L_4$ or $L_5$ produces $\msome$ in $\ssome$ with probability $S_{0}(\msome \mid
\ssome;L_{4,5}) \approx .73$. The probability of producing $\msome$ in $\sall$ is $S_{0}(\msome \mid
\sall;L_4) \approx .27$ for $L_4$ and $S_{0}(\msome \mid
\sall;L_4) \approx .5$ for $L_5$. \mf{I realize that the probability of producing a false
  message for literal speakers is actually rather high. That's iffy because our pragmatic types
are able (given favorable parameter values) to perform much better. That means that we
implicitly have engineered in a possible fitness advantage of pragmatic types! Should we do
something about this? Use definition $S_{0}(m \mid s;L) \propto L_{sm} + 0.001$? Make sure we
only look at parameters for pragmatic types that make them not much better than the literal guys?}

Pragmatic behavior of level-$n+1$ is similar to its literal counterparts but uses the
interpretation or production behavior of a level-$n$ player instead of the lexical meaning:
\begin{flalign}
&H_{n+1}(s|m;L) \propto pr(s) S_{n}(m|s;L) \label{h:leveln}\\
&S_{n+1}(m|s;L) \propto  \exp(\lambda \; H_{n}(s|m;L)^\alpha) \label{s:leveln}
\end{flalign}
As usual in probabilistic pragmatics models, speaker behavior is regulated by a soft-max
parameter $\lambda$, $\lambda \geq 1$ \citep{luce:1959,sutton+barto:1998}. As $\lambda$
increases, choices made in production are more rational in that higher values lead to behavior
that is increasingly in line with expected utility maximization. Expected utility of a message
$\messg$ in state $\state$ is here defined as $H_{n}(s|m;L)^\alpha$. The probability
$H_{n}(s|m;L)$ is the probability that the hearer will assign to or choose the correct
meaning. The transformation with $\alpha$ allows to differentiate theoretically interesting
speaker production behavior. To see this, notice that for two states and two messages (the
minimal scenario we will investigate), the possible values for $H_{0}(s|m;L)$ are restricted to
zero, $0.5$ and one. Consider the case of $H_0$ with $L_5$, which assigns the following
probabilities to states after hearing messages:
\begin{align*}
  H_{0}(\messg \mid \state; L_5) & = \begin{blockarray}{lcc}
    & \mygray{\ssome} & \mygray{\sall} \\
    \begin{block}{l[cc]}
      \mygray{\msome} & 0.5 & 0.5 \\
      \mygray{\mall}  & 0 & 1 \\
    \end{block}
  \end{blockarray}
\end{align*}
A level-1 speaker's utilities are then:
\begin{align*}
  \begin{blockarray}{lcc}
    & \mygray{\msome} & \mygray{\mall} \\
    \begin{block}{l[cc]}
      \mygray{\ssome} & 0.5^\alpha & 0^\alpha \\
      \mygray{\sall}  & 0.5^\alpha & 1^\alpha \\
    \end{block}
  \end{blockarray}
\end{align*}
As the soft-max choice rule is sensitive to differences in utilities, no matter what $\lambda$
is we will get $S_1(\msome \mid \ssome; L_5) = S_1(\mall \mid \sall; L_5)$ if $\alpha =
1$. This means that we cannot distinguish pragmatically deviant choices (selecting a less
informative message instead of a more informative one), from semantically deviant choices
(selecting a false message instead of a true one). If $\alpha > 1$ pragmatically deviant
choices will get less likely than semantically deviant choices; if $\alpha < 0$ it is the other
way around. In sum, parameter $\lambda$ regulates the overall tendency to avoid pragmatically
or semantically deviant choices, $\alpha$ further distinguishes between pragmatically and
semantically deviant behavior. Variation in $\alpha$ is only necessary for the technical
extension to environmental transmission noise in Section~XYZ. \mf{I would suggest to just give
  the intuitive summary here and have a little appendix paragraph with the example. What do you
say?}

>>>>

\definecolor{Red}{RGB}{178,34,34}
\newcommand{\mf}[1]{\textcolor{Red}{[MF: #1]}} 
\newcommand{\tb}[1]{\textcolor[rgb]{.8,.33,.0}{[TB: #1]}}% prints in orange

\usepackage{blkarray}
\usepackage{xspace}

@@

A particularly productive and well studied class of pragmatic enrichments are scalar
implicatures
\citep{horn:1984,Hirschberg1985:A-Theory-of-Sca,LevinsonPragmatics1983,Geurts2010:Quantity-Implic}. Usually,
an utterance of a sentence like ``I own some of Johnny Cash's albums'' will be taken to mean
that the speaker does not own all of them. This is because, if the speaker had them all, he
could have used the stronger word \emph{all} instead of \emph{some} in his utterance, and
thereby would have made a more informative statement. Scalar implicatures, especially the
inference from \emph{some} to \emph{some but not all}, have been studied extensively, both
theoretically
\citep[e.g.][]{Sauerland2004:Scalar-Implicat,ChierchiaFox2008:The-Grammatical,Rooyvan-RooijJagerde-Jager2012:Explaining-Quan}
as well as experimentally
\citep[e.g.][]{BottNoveck2004:Some-Utterances,huang+snedeker:2009,GrodnerKlein2010:Some-and-Possib,GoodmanStuhlmuller2013:Knowledge-and-I,DegenTanenhaus2012:Processing-Scal}. While
there is much dispute in this domain about many interesting details, a position endorsed by a
clear majority is that a word like \emph{some} is underspecified to mean \emph{some and maybe
  all} and that the enrichment to \emph{some but not all} is part of some regular 
process with roots in pragmatics.

If this majority view is correct, the question arises how such a division of labor between
semantics and pragmatics could have evolved. Models of language evolution abound. There are
simulation-based models studying the evolution of language in populations of communicating
agents
\citep{Hurford1989:Biological-Evol,Steels1995:A-Self-Organizi,LenaertsJansen2005:The-Evolutionar,SteelsBelpaeme2005:Coordinating-Pe,BaronchelliPuglisi2008:Cultural-route-,steels:2011,SpikeStadler2016:Minimal-Require}
and there are mathematical models of language evolution, mostly coming from game theory
\citep{lewis:1969,Warneryd1993:Cheap-Talk-Coor,BlumeKim1993:Evolutionary-St,nowak+krakauer:1999,Huttegger2007:Evolution-and-t,Skyrms2010:Signals}. Much
work has focused on explaining basic properties such as compositionality and combinatoriality
\citep[e.g.][]{Batali1998:Computational-S,nowak+krakauer:1999,nowak+etal:2000,KirbyHurford2002:The-Emergence-o,kirby:2002,SmithKirby2003:Iterated-Learni,Gong2007:Language-Evolut,kirby+etal:2015,verhoef+etal:2014,Franke2015:Proto-Syntax}, but
little attention has been paid to the interaction between conventional meaning and pragmatic
use. What is more, many mathematical models explain evolved meaning as a regularity in the
behavior of agents which maps objective states of the world to observable signals. There is no
room in such a purely extensional approach to address the semantics-pragmatics division
directly. Instead, we would need to look at richer representations of cognizing agents and their
communicative interaction.

To fill this gap, we spell out a model of the co-evolution of conventional meaning and
pragmatic reasoning types. The objects of replication and selection are pairs of lexical
meanings and general types of pragmatic behavior, which we represent using state-of-the-art
probabilistic cognitive models of pragmatic language use
\citep{frank+goodman:2012,FrankeJager2015:Probabilistic-p,GoodmanFrank2016:Pragmatic-Langu}. Replication
and selection are described by the \emph{replicator mutator dynamic}, a general and established
model of evolutionary change in large and homogeneous populations
\citep{Hofbauer1985:The-Selection-M,nowak+etal:2000,NowakKomarova2001:Evolution-of-Un,hofbauer+sigmund:2003,Nowak2006:Evolutionary-Dy}. The
approach allows us to study the interaction between (i) evolutionary pressure towards
communicative efficiency and (ii) possible infidelity in the transmission of linguistic
knowledge, caused by factors such as inductive learning biases or systematic perceptual errors. Considering
transmission of linguistic knowledge is important because neither semantic meanings nor
pragmatic usage patterns are directly observable. Instead, language learners have to infer these
unobservables from the observable behavior in which they result. We formalize this 
process as a form of Bayesian inference. Our approach thereby contains a well-understood model
of iterated Bayesian learning \citep{griffiths+kalish:2007} as a special case, but combines it
with functional selection, here formalized as the most versatile dynamic from evolutionary game
theory; the replicator dynamic
\citep{TaylorJonker1978:Evolutionary-St}. Section~\ref{sec:model} introduces this model.

Section~\ref{sec:si-case-study} applies this model to a case study on scalar implicatures. We
discuss a setting in which the majority view of underspecified lexical meanings and pragmatic
enrichments emerges if selection and transmission infidelity are combined. In particular, we
show that inductive learning biases of Bayesian learners that favor simpler lexical meanings
can lead to this outcome. Additionally, we show that a similar outcome can be reached 
without assuming any cognitive biases, but simply as an epiphenomenon of systematic disturbances from
environmental factors. This formal results highlights the frequently overlooked possibility
that channel noise in evolutionary replication can mimic effects of inductive biases.

We see the main contribution of this work as conceptual and technical, not as a definite answer
to the question why scalar implicatures emerge. The work here rather demonstrates how current
probabilistic cognitive modeling of language use and evolutionary modeling can be fruitfully
combined to study the co-evolution of semantics and pragmatics side-by-side. Reversely, the
approach taken here may be seen as a first step towards giving an evolutionary rationale for
empirically successful probabilistic models of language use that embrace the majority view of
the division of labor between semantics and pragmatics. Section~\ref{sec:discussion} elaborates
on these points.

@@

The emergence and change of linguistic structure is influenced by many intertwined
factors. These range from biological and socio-ecological to cultural ones \citep{benz+etal:2005b,steels:2011,tamariz+kirby:2016}. Social and ecological pressures determine communicative needs, while
biology determines the architecture that enables and constrains the means by which they can be
fulfilled. In the following, our focus lies on the cultural aspects, wherein processes of
linguistic change are viewed as shaped by language use and its transmission, i.e., as a result
of a process of cultural evolution
\citep{Pagel2009:Human-Language-,ThompsonKirby2016:Culture-Shapes-}.

The idea that language is an adaptation to serve a communicative function has played a pivotal
role in synchronic and diachronic analyses at least since \citeposs{zipf:1949} explanation of
word frequency rankings as a result of competing hearer and speaker preferences (e.g. in
\citealt{martinet:1962, horn:1984,jaeger+vRooij:2007,jaeger:2007,
  piantadosi:2014,kirby+etal:2015}). If processes of selection, such as conditional imitation
or reinforcement, favor more communicatively efficient types of behavior, languages are driven
towards semantic expressivity \citep[e.g.][]{nowak+krakauer:1999,Skyrms2010:Signals}. But
pressure towards communicative efficiency is not the only force that shapes
language. Learnability is another as natural languages need to be learnable to survive their
faithful transmission across generations. Clearly, an unlearnable code will not make it past
the one happy fellow who invented it. Importantly, even small biases implicit in acquisition
can build up and have quite striking effects on an evolving language in a process
of iterated learning
\citep{KirbyHurford2002:The-Emergence-o,SmithKirby2003:Iterated-Learni,kirby+etal:2014}.

While natural languages are pressured for both expressivity and learnability these forces may pull in opposite directions. The opposition becomes particularly clear when
considering the extreme (cf. \citealt{kemp+regier:2012,kirby+etal:2015}). A language with a
single form-meaning association is easy to learn but lacking in expressivity. Conversely, a
language that associates a distinct form with all possible meanings a speaker may want to
convey is maximally expressive but challenging to acquire.

An elegant formal approach to capturing the interaction between expressivity and learnability
is the \emph{replicator mutator dynamic}
\citep{Hofbauer1985:The-Selection-M,nowak+etal:2000,NowakKomarova2001:Evolution-of-Un,hofbauer+sigmund:2003,Nowak2006:Evolutionary-Dy}. In
its simplest, discrete-time formulation, the RMD defines the frequency $x'_i$ of each type $i$
in a population at the next time step as a function of: (i) the frequency $x_i$ of each type
$i$ before the update, (ii) the fitness $f_i$ of each type $i$ before the update, and (iii) the
probability $Q_{ji}$ that an agent who wants to imitate, adopt, or learn the type of an agent
with type $j$ ends up acquiring type $i$:
\begin{align}
  \label{eq:RMD_discrete}
  x'_i = \sum_j Q_{ji} \frac{x_jf_j}{\sum_k x_k f_k}\,.
\end{align}
The RMD consists of two components: fitness-based selection and transmission biases. This
becomes most transparent when we consider an equivalent formulation in terms of a step-wise
application of the discrete-time replicator dynamic \citep{TaylorJonker1978:Evolutionary-St} on the initial population vector $\vec{x}$
and its subsequent multiplication with a mutation matrix $Q$:

\begin{align}
  \label{eq:RMD_discrete_recast}
  x'_i & = (\text{M}(\text{RD}(\vec{x})))_i\,,
\end{align}
where
\begin{align*}
      \left ( \text{RD}(\vec{x}) \right )_i 
         = \frac{x_i f_i}{\sum_k x_k f_k}
 \ \ \ \ \text{and} \ \ \ \ 
  (\text{M}(\vec{x}))_i = (\vec{x} \cdot Q)_i = \left ( \sum_j
          x_j Q_{ji} \right)_i\,.
\end{align*}
If the transmission matrix $Q$ is trivial in the sense that $Q_{ji}=1$ whenever $j=i$, the
dynamic reduces to the replicator dynamic. The replicator dynamic is a model of fitness-based
selection in which the relative frequency of type $i$ will increase with a gradient
proportional to its average fitness in the population. This dynamic is popular and
versatile because it can be derived from many abstract processes of biological and cultural
transmission and selection \citep[for overview and several derivations
see][]{Sandholm2010:Population-Game}, including conditional imitation
\citep[e.g.][]{Helbing1996:A-Stochastic-Be,Schlag1998:Why-Imitate-and} or reinforcement
learning \citep[e.g.][]{BorgersSarin997:Learning-Throug,Beggs2005:On-the-Converge}. If fitness
$f_i$ is the same for all types $i$, the replicator step is the identity map
$ \left ( \text{RD}(\vec{x}) \right )_i = x_i$ and the dynamic reduces to a process of
iteration of the transmission bias encoded in $Q$. In this way, the process in
(\ref{eq:RMD_discrete}), equivalently (\ref{eq:RMD_discrete_recast}), contains a model of
iterated learning \citep{griffiths+kalish:2007}. \mf{should we include a simple example here? I 
have an example from a lecture ready at hand; it's a simple coordination game in a
one-population setting.} \tb{Yes, I think most readers would appreciate it. We can always cut it out again if the reviewers prefer more compression.}

@@

Types are what cultural evolution operates on. In standard applications of evolutionary game
theory, types correspond to ways of acting in a game, e.g., either cooperating or defecting in
a prisoner's dilemma. \mf{maybe good to refer back to the example from before if there was
  one?} \tb{Yes!} For our purposes here, types are identified by their cognitive make-up. Since we are
interested in the question under which conditions processes of cultural evolution will favor
specific divisions of labor between lexical meaning and pragmatic use, a type is a pair
consisting of a lexicon and a pragmatic strategy.

Lexica codify the truth-conditions of expressions. A convenient way to represent lexica is by
$(\card{\States}, \card{\Messgs})$-Boolean matrices, where $\States$ is a set of states
(meanings) and $M$ a set of messages (forms available in the language). For example, suppose
that there are two relevant world states $\States = \set{\ssome, \sall}$. In state $\ssome$
Chris owns some but not all of Johnny Cash's albums while in $\sall$ Chris owns them
all. Suppose that there are two messages $\Messgs = \set{\msome, \mall}$ where $\msome$ is
short for a sentence like \emph{Chris owns some of Johnny Cash's albums} and $\mall$ for the
same sentence with \emph{some} replaced by \emph{all}. Lexica for this case would assign a
truth value for each state-message pair. The following two lexica exemplify the distinction
between a lexicalized upper-bound for \emph{some} in $L_4$ and the widely assumed logical
semantics with only a lower-bound in $L_5$. \mf{can we not give these guys more mnemonic
  names?} \tb{``bound'' and ``lack''? They would need to be pretty short to not be too disruptive as an inline indices.}
\begin{align*}
  L_{4} & = \begin{blockarray}{lcc}
    & \mygray{\msome} & \mygray{\mall} \\
    \begin{block}{l[cc]}
      \mygray{\ssome} & 1 & 0 \\
      \mygray{\sall}  & 0 & 1 \\
    \end{block}
  \end{blockarray} &
  L_{5} & = \begin{blockarray}{lcc}
    & \mygray{\msome} & \mygray{\mall} \\
    \begin{block}{l[cc]}
      \mygray{\ssome} & 1 & 0 \\
      \mygray{\sall}  & 1 & 1 \\
    \end{block}
  \end{blockarray}
\end{align*}

We distinguish between two kinds of pragmatic behavior. {\em Literal interlocutors} produce and
interpret messages literally, being guided only by their lexica. {\em Pragmatic interlocutors}
instead engage in mutual reasoning to inform their choices. Recent probabilistic models of
rational language use
\citep{franke:2009,frank+goodman:2012,FrankeJager2015:Probabilistic-p,GoodmanFrank2016:Pragmatic-Langu}
capture different types of pragmatic behavior in a reasoning hierarchy. The hierarchy's bottom,
level $0$, corresponds to literal language use. Pragmatic language users of level $n + 1$
act (approximately) rational with respect to level-$n$ behavior of their
interlocutors. (\ref{h:level0}) and (\ref{s:level0}) define probabilistic behavior of literal
hearers and speakers respectively: \mf{is it a problem that $S$ denotes speakers and the set of
states?} \tb{I think that's fine. $S_n(\cdot \mid \cdot)$ is not used beyond this section and the set $S$ only by mention of its members. Otherwise we can, e.g., relabel $M$ as $F$(orms) and use $M$(eanings) instead of $S$(tates). I'm fine with either alternative.}
\begin{flalign}
&H_{0}(s \mid m;L) \propto pr(s) L_{sm} \label{h:level0}\\
&S_{0}(m \mid s;L) \propto \exp(L_{sm}) \label{s:level0}
\end{flalign}
According to (\ref{h:level0}), a literal hearer's interpretation of a message $m$ as a state
$s$ depends on her lexicon and her prior over states, $pr \in \Delta(S)$. For simplicity, in
the following this prior is assumed to be uniform. A literal interpreter with lexicon $L_4$
from above would assign $\ssome$ a probability of $H_0(\ssome \mid \msome; L_4) = 1$ after
hearing $\msome$, while a literal interpreter with lexicon $L_5$ would assign $\ssome$
probability $H_0(\ssome \mid \msome; L_5) = 0.5$. A literal speaker chooses any true message
for a state with equal probability and prefers a true message over a false one. The formulation
in (\ref{s:level0}) also allows false messages to be sent with a small positive probability in
analogy to the definition of pragmatic speakers in probabilistic pragmatic models. This is
necessary to guarantee a mutation matrix with only positive entries (see below). A literal
speaker with either lexicon $L_4$ or $L_5$ produces $\msome$ in $\ssome$ with probability $S_{0}(\msome \mid
\ssome;L_{4,5}) \approx .73$. The probability of producing $\msome$ in $\sall$ is $S_{0}(\msome \mid
\sall;L_4) \approx .27$ for $L_4$ and $S_{0}(\msome \mid
\sall;L_5) \approx .5$ for $L_5$. \mf{I realize that the probability of producing a false
  message for literal speakers is actually rather high. That's iffy because our pragmatic types
are able (given favorable parameter values) to perform much better. That means that we
implicitly have engineered in a possible fitness advantage of pragmatic types! Should we do
something about this? Use definition $S_{0}(m \mid s;L) \propto L_{sm} + 0.001$? Make sure we
only look at parameters for pragmatic types that make them not much better than the literal guys?} \tb{That's true. We could also just let them soft-maximize. Let's decide on this via email. I'll adapt the simulation results from the next section according to what we decide.}

Pragmatic behavior of level-$n+1$ is similar to its literal counterparts but uses the
interpretation or production behavior of a level-$n$ player instead of the lexical meaning:
\begin{flalign}
&H_{n+1}(s|m;L) \propto pr(s) S_{n}(m|s;L) \label{h:leveln}\\
&S_{n+1}(m|s;L) \propto  \exp(\lambda \; H_{n}(s|m;L)^\alpha) \label{s:leveln}
\end{flalign}
As usual in probabilistic pragmatics models, speaker behavior is regulated by a soft-max
parameter $\lambda$, $\lambda \geq 1$ \citep{luce:1959,sutton+barto:1998}. As $\lambda$
increases, choices made in production are more rational in that higher values lead to behavior
that is increasingly in line with expected utility maximization. Expected utility of a message
$\messg$ in state $\state$ is here defined as $H_{n}(s|m;L)^\alpha$. The probability
$H_{n}(s|m;L)$ is the probability that the hearer will assign to or choose the correct
meaning. The transformation with $\alpha$ allows to differentiate theoretically interesting
speaker production behavior: If $\alpha > 1$ pragmatically deviant
choices will get less likely than semantically deviant choices; if $\alpha < 0$ it is the other
way around. See Appendix \ref{app:sig-example} for an example that elucidates the role of $\alpha$ \tb{Did you mean it like this?}.

In sum, parameter $\lambda$ regulates the overall tendency to avoid pragmatically or semantically deviant choices, $\alpha$ further distinguishes between pragmatically and
semantically deviant behavior. Variation in $\alpha$ is only necessary for the technical
extension to environmental transmission noise in Section \ref{subsec:noise}.
